A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-10-06 inst/paper.Rnw
warning added in case not all vertex information is included

\hh{todo list:
\begin{enumerate}
\item include installation instruction for package ggnet
\item the biggest change to the previous implementation is that there is no longer a vertices data set. Instead, all of the vertex information has to be included in the edge dataset. The assumption is that any vertex information included relates to the `from\_id' column. If the user does not include ALL vertex information in the dataset (i.e. if some vertices only show up under the `to\_id' column), there will be segment ends without a dot in the graphic. As a measure against that, we suggest to merge edges and vertex information along the `from\_id' column using the parameter setting `all = TRUE'. This ensures all vertices to be included; if a vertex is a sink, the corresponding `to\_id' value is set to NA. This is accommodated for both in the layout and the drawing of the graph.
\item the parameters have changed names: all of the vertex parameters have the regular names, edge parameters start with `e'. `esize' is re-named to `linewidth'.
Some parameters might not be implemented yet, we need to get a good overview of what is missing and what works.
\item we could draw selfies, but at the moment we don't. We need to discuss this.
\end{enumerate}
}
\tableofcontents
\newpage
\section{Introduction}
At its core, a network is simply a set of points connected in pairs by a set of lines \citep{newman}.  Here, we refer to the lines as edges and the points as vertices, although these are also called nodes.  These two seemingly simple sets of graphical objects, points and segments, are used to encode a huge variety and quantity of information across many fields of study. For instance, networks of scientific collaboration, a food web of marine animals, and American college football games are all covered in a paper on community detection in networks by \citet{football}.  Additionally, \citet{networkfailures} examine node failure in interdependent networks like power grids.  Social networks, such as links between actors found on \url{www.imdb.com}, and neural networks, like the completely mapped neural network of the \textit{C. elegans} worm are also etensively studied \citep{smallworld}.  Networks vary widely in scope and complexity: the smallest network is simply an edge between two vertices, while one of the most commonly used and most complex networks, the world wide web, has billions of vertices (webpages) and billions of edges (hyperlinks) connecting them.  The edges in a network can  be directed or undirected: directed edges represent information travelling from one vertex to another, and switching the direction would change the structure of the network. The world wide web is an example of a directed network because one webpage may link to another, and not necessarily the other way around. Undirected edges, however, are simply connections between vertices.  Coauthorship networks that encode information about academic publications are examples of undirected networks because if two people author a paper together, that creates a connection between them that is bidirectional.

@@

  \subsection{Data Structure}
In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. The only necessary variable is one called \texttt{label} which contains the vertex labels as they are encoded in the edges data set. This will be a formal requirement in the final version of the \texttt{geom\_net} function.  Each row represents one vertex.  Other values of interest, such as grouping variables or degree of each vertex should be stored as columns in this dataframe, with an observation for each vertex.
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should match the names of the vertices in the vertex information data frame. The formal names for these columns are \texttt{from\_id} and \texttt{to\_id}, respectively.  Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.

@@

\section{Usage Examples}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} in a series of diverse examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.
<<blood_ex, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.", fig.width=6, fig.height=6>>=
ggplot(data = bloodnet, aes(from_id = from, to_id = to)) +
  geom_net(colour="red", layout="circle", label=TRUE, size=3,
           directed=TRUE, vjust=-0.5, arrowsize=1.5) + theme_net()
@
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{vcolour} and \texttt{vsize} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{vlabel} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has out degree of seven and in degree of zero. Additionally, you can see that the AB+ type is the universal recepient, with in degree of seven and out degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figure/blood_ex-1.pdf}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}

>>>>

\hh{todo list:
\begin{enumerate}
\item change format to comply with journal
\item include installation instruction for package ggnet
\item the biggest change to the previous implementation is that there is no longer a vertices data set. Instead, all of the vertex information has to be included in the edge dataset. The assumption is that any vertex information included relates to the `from\_id' column. If the user does not include ALL vertex information in the dataset (i.e. if some vertices only show up under the `to\_id' column), there will be segment ends without a dot in the graphic. As a measure against that, we suggest to merge edges and vertex information along the `from\_id' column using the parameter setting `all = TRUE'. This ensures all vertices to be included; if a vertex is a sink, the corresponding `to\_id' value is set to NA. This is accommodated for both in the layout and the drawing of the graph.
\item the parameters have changed names: all of the vertex parameters have the regular names, edge parameters start with `e'. `esize' is re-named to `linewidth'.
Some parameters might not be implemented yet, we need to get a good overview of what is missing and what works.
\item we could draw selfies, but at the moment we don't. We need to discuss this.
\end{enumerate}
}
\tableofcontents
\newpage
\section{Introduction}
At its core, a network is simply a set of points connected in pairs by a set of lines \citep{newman}.  Here, we refer to the lines as edges and the points as vertices, although these are also called nodes.  These two seemingly simple sets of graphical objects, points and segments, are used to encode a huge variety and quantity of information across many fields of study. For instance, networks of scientific collaboration, a food web of marine animals, and American college football games are all covered in a paper on community detection in networks by \citet{football}.  Additionally, \citet{networkfailures} examine node failure in interdependent networks like power grids.  Social networks, such as links between actors found on \url{www.imdb.com}, and neural networks, like the completely mapped neural network of the \textit{C. elegans} worm are also etensively studied \citep{smallworld}.  Networks vary widely in scope and complexity: the smallest network is simply an edge between two vertices, while one of the most commonly used and most complex networks, the world wide web, has billions of vertices (webpages) and billions of edges (hyperlinks) connecting them.  The edges in a network can  be directed or undirected: directed edges represent information travelling from one vertex to another, and switching the direction would change the structure of the network. The world wide web is an example of a directed network because one webpage may link to another, and not necessarily the other way around. Undirected edges, however, are simply connections between vertices.  Coauthorship networks that encode information about academic publications are examples of undirected networks because if two people author a paper together, that creates a connection between them that is bidirectional.

@@

\section{Our Implementation}
\hh{this section needs a better name, but we need it, because here is where your work starts. it should not be part of the introduction.}
  \subsection{Data Structure}
In order for this geometry to work using data frames, there need to be two separate data frames given to the \texttt{geom\_net} function: one for the edge information and one for the vertex information. The vertex data frame should contain all the relevant vertex information. The only necessary variable is one called \texttt{label} which contains the vertex labels as they are encoded in the edges data set. This will be a formal requirement in the final version of the \texttt{geom\_net} function.  Each row represents one vertex.  Other values of interest, such as grouping variables or degree of each vertex should be stored as columns in this dataframe, with an observation for each vertex.
  \par The edge data frame should contain all the relevant edge information. The only necessary variables are the ``from vertex" and ``to vertex" for each edge in the network. The From and To vertices should match the names of the vertices in the vertex information data frame. The formal names for these columns are \texttt{from\_id} and \texttt{to\_id}, respectively.  Other variables may also be included for each edge, such as the edge weight or grouping variable.  As before, the variables of interest are columns in the data frame and the rows are each edge in the network.

@@

\section{Examples of Networks}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} in a series of diverse examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.
<<blood_ex, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.", fig.width=6, fig.height=6>>=
ggplot(data = bloodnet, aes(from_id = from, to_id = to)) +
  geom_net(colour="red", layout="circle", label=TRUE, size=3,
           directed=TRUE, vjust=-0.5, arrowsize=1.5) + theme_net()
@
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{vcolour} and \texttt{vsize} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{vlabel} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has out degree of seven and in degree of zero. Additionally, you can see that the AB+ type is the universal recepient, with in degree of seven and out degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figure/blood_ex-1.pdf}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}
